OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-11-22 manuscript_nar.tex
adding supp figures and fixing abstract
Adding @jaquol comments

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 2~kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined by
\cite{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
\citet{yan2017hicspector} sums the distances between the leading twenty
eigenvectors of the Laplacian of the Hi-C matrix. This approach borrows
the concepts of spectral clustering \citep{von2007tutorial} and amounts to
comparing high level features of the matrix.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_1.eps}}
\caption{Modeling biases from the contact profile. A. Non linear
relationship between the number of restriction sites and the total number
of contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. The brown line represents the raw signal, the blue
line represents the signal after bias correction, the black line
represents the signal after bias and copy number (CN) correction. The long
arm of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
four copies, explaining that the signal is about twice higher than for the
short arm.}
\label{fig:totals}
\end{figure}

@@

We collected multiple Hi-C data sets obtained from T47D and K562 cells,
both with aberrant karyotypes. The first pool, referred to as the T47D data
set contained six T47D samples and two K562 samples, the second, refered
to as the K562 data set contained eight T47D samples and two T47D samples.
We compared matrices before and after normalization by different methods
(see Materials and Methods, distributions shown in Supplementary Figures~2
and 3). This gave an indication of the impact of a given normalization
method on the biological variation (differences between samples from T47D
and K562) and on the technical variation (differences among samples from
the same cell type). The results are summarized in
Figure~\ref{fig:aberrant}.

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_2.eps}}
\caption{Removing biases from Hi-C on aberrant karyotypes. A and B.
Average changes compared to the raw data on the T47D and K562 data sets.
The bars represent 95\% confidence intervals centered on the mean
difference of the reproducibility score between a given correction method
and the raw data. C and D. ROC curves on the T47D and K562 sets. The areas
under the curve are indicated in the bottom right corner. The color code
is the same as in panels A and B.}
\label{fig:aberrant}
\end{figure}

@@

As above, we used the reproducibility score for classification and
compared the tools with a ROC curve (Figure~\ref{fig:normal}B). The
performance of the all tools were high and less variable than for
aneuploid cell lines. \textit{OneD} achieved the highest area under the
curve on this problem, but with a small margin over \textit{KR} and
\textit{ICE}. The performance of \textit{OneD+CN} was the same as that of
\textit{OneD} because copy numer correction has no effect on diploid cell
lines. Using other metrics to compare matrices gave similar results
(Supplementary Figures~8, 9 and 10).

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_3.eps}}
\caption{Removing biases from Hi-C on normal karyotypes. A. Average
changes compared to the raw data on the mouse data set.  B. ROC curves on
the mouse data set. The legend and color code are as in
Figure~\ref{fig:aberrant}.}
\label{fig:normal}
\end{figure}

>>>>

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 5~kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mbp in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined by
\cite{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
\citet{yan2017hicspector} sums the distances between the leading twenty
eigenvectors of the Laplacian of the Hi-C matrix. This approach borrows
the concepts of spectral clustering \citep{von2007tutorial} and amounts to
comparing high level features of the matrix.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_1.eps}}
\caption{Modeling biases from the contact profile. A. Non linear
relationship between the number of restriction sites and the total number
of contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. The purple line represents the raw signal, the dark blue
line represents the signal after OneD bias correction, the light blue line
represents the signal after OneD bias and copy number (CN) correction and
the orange line represents the signal after ICE bias correction. The long arm
of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
four copies, explaining that the signal is about twice higher than for
the short arm.}
\label{fig:totals}
\end{figure}

@@

We collected multiple Hi-C data sets obtained from T47D and K562 cells,
both with aberrant karyotypes. The first pool, referred to as the T47D data
set contained six T47D samples and two K562 samples, the second, refered
to as the K562 data set contained eight K562 samples and two T47D samples.
We compared matrices before and after normalization by different methods
(see Materials and Methods, distributions shown in Supplementary Figures~2
and 3). This gave an indication of the impact of a given normalization
method on the biological variation (differences between samples from T47D
and K562) and on the technical variation (differences among samples from
the same cell type). The results are summarized in
Figure~\ref{fig:aberrant}.

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_2.eps}}
\caption{Removing biases from Hi-C on aberrant karyotypes. A and B.
Average changes compared to the raw data on the T47D and K562 data sets.
The bars represent 95\% confidence intervals centered on the mean
difference of the reproducibility score between a given correction method
and the raw data. Upper left panels: Different cell type samples
processed using different protocols. Upper right panels: Same cell
type samples processed using different protocols. Lower left panels:
Same cell type samples processed using the same protocol. C and D. ROC
curves on the T47D and K562 sets. The areas under the curve are
indicated in the bottom right corner. The color code is the same as in
panels A and B.}
\label{fig:aberrant}
\end{figure}

@@

As above, we used the reproducibility score for classification and
compared the tools with a ROC curve (Figure~\ref{fig:normal}B). The
performance of all the tools were high and less variable than for
aneuploid cell lines. \textit{OneD} achieved the highest area under the
curve on this problem, but with a small margin over \textit{KR} and
\textit{ICE}. The performance of \textit{OneD+CN} was the same as that of
\textit{OneD} because copy numer correction has no effect on diploid cell
lines. Using other metrics to compare matrices gave similar results
(Supplementary Figures~8, 9 and 10).

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_3.eps}}
\caption{Removing biases from Hi-C on normal karyotypes. A. Average
  changes compared to the raw data on the mouse data set.
  Upper left panel: Different cell type samples
processed using different protocols. Upper right panel: Same cell
type samples processed using different protocols. Lower left panel:
Same cell type samples processed using the same protocol. Lower right panel:
Different cell type samples processed using the same protocol. B. ROC curves on
the mouse data set. The legend and color code are as in Figure~\ref{fig:aberrant}.}
\label{fig:normal}
\end{figure}
